---
title: "phoenix: An R package and Python module for calculating the Phoenix Pediatric Sepsis Score and Criteria"

code-fold: false
echo: true
number-sections: false

format:
  html:
    fig-cap-location: top
  docx:
    fig-cap-location: top
    reference-doc: template.docx

bibliography: references.bib
csl: jamia.csl
---

```{r}
#| label: setup
#| include: false
source("setup.R")
```


**Authors**: Peter E. DeWitt^1^, PhD; Seth Russell^1^, MS; Margaret N. Rebull^1^, MA; L. Nelson Sanchez-Pinto^2^, MD, MBI; Tellen D. Bennett^1,3^, MD, MS

[Author affiliations:]{.underline}

^1^ Department of Biomedical Informatics, University of Colorado School of Medicine, University of Colorado Anschutz Medical Campus, Aurora, Colorado, USA

^2^ Departments of Pediatrics (Critical Care) and Preventive Medicine (Health and Biomedical Informatics), Northwestern University Feinberg School of Medicine, and Ann and Robert H. Lurie Children’s Hospital of Chicago, Chicago, Illinois, USA

^3^ Section of Critical Care Medicine, Department of Pediatrics, University of Colorado School of Medicine, University of Colorado Anschutz Medical Campus, Aurora, Colorado, USA

[**Correspondence:**]{.underline}

Peter E. DeWitt, PhD

Department of Biomedical Informatics

University of Colorado School of Medicine

1890 N. Revere Court

Mailstop F600

Aurora, Colorado, 80045

Email: [peter.dewitt@cuanschutz.edu](mailto:peter.dewitt@cuanschutz.edu)

[**Article type:**]{.underline} Application Note

[**Word count:**]{.underline} 1787 / 2000 (excluding tables)

[**Tables:**]{.underline} 2

[**Figures:**]{.underline} 0

[**Supplementary material:**]{.underline} Supplementary Examples

[**Short running title**]{.underline}: R package and Python module for calculating the Phoenix criteria

<!-- allowed 5; recomend https://www.ncbi.nlm.nih.gov/mesh/ -->
[**Keywords:**]{.underline} Sepsis, Septic shock, pediatrics, computer software, EHR

{{< pagebreak >}}

# Abstract {.unnumbered}

words: 145/150 words

**Objectives**

The publication of the Phoenix criteria for pediatric sepsis and septic shock
initiates a new era in clinical care and research of pediatric sepsis. Tools to
consistently and accurately apply the Phoenix criteria to electronic health
records (EHRs) is one part of building a robust and internally consistent body
of research across multiple research groups and data sets.

**Materials and Methods**

We developed the `phoenix` R package and Python module to provide researchers
with intuitive and simple functions to apply the Phoenix criteria to EHR data.

**Results**

The `phoenix` R package and Python module enable researchers to apply the
Phoenix criteria to EHR datasets and derive the relevant indicators, total
scores, and sub-scores.

**Conclusions**

The `phoenix` R package and Python model are freely available on CRAN, PyPi, and
GitHub. These tools enable the consistent and accurate application of the
Phoenix criteria to EHR datasets.


{{< pagebreak >}}

::: {.callout-tip icon="false"}
## Lay Summary
<!-- word count: 154 / 200 -->

The Phoenix criteria for pediatric sepsis and septic shock mark the beginning of
a new era for clinical care and research of pediatric sepsis. The new
criteria represent a conceptual shift in defining sepsis, moving away from an
inflammatory based criteria to life-threatening organ dysfunction based
criteria. As these new diagnostic criteria move to the bedside and research
areas, the need to consistently and accurately apply the Phoenix criteria to
electronic health records (EHR) data is necessary to build a robust and
internally consistent body of research.

We developed the `phoenix` R package and Python module. These two freely
available tools, along with example SQL queries, provide pediatric sepsis
researchers the means to apply the Phoenix criteria to EHR data without
needing to implement the criteria themselves. Use of `phoenix` will reduce
sources of error in research and provide a common implementation for multiple
research groups, across multiple data formats, and multiple programming
paradigms.

:::

{{< pagebreak >}}

# Background and Significance

Approximately 3.3 million pediatric deaths per year are attributable to sepsis
and septic shock.[@rudd_2020_global]  In January 2024, the Phoenix diagnostic
criteria for pediatric sepsis were published to supersede the
criteria[@goldstein_2005_international] defined by the International Pediatrics
Sepsis Criteria Conference in
2005.[@schlapbach_2024_international;@sanchezpinto_2024_development]
Transitioning to the Phoenix criteria is a conceptual change, moving away from
an inflammatory response-based criteria to a life-threatening organ
dysfunction-based criteria.  This change parallels the conceptual change for the
diagnostic criteria of adult sepsis.[@seymour2016assessment;@singer2016third]

The Pediatric Sepsis Task Force developed the Phoenix criteria using a
data-driven modified Delphi consensus approach. The criteria are based on four
organ dysfunctions: respiratory, cardiovascular, coagulation, and neurologic.
Additionally, the task force published an 8-organ system score, Phoenix-8, for
research
purposes.[@sanchezpinto_2024_development;@schlapbach_2024_international]

The Phoenix criteria initiate a new era of benchmarking,
epidemiological surveillance, clinical quality improvement, and research in
pediatric sepsis.

Sufficient information to implement the Phoenix criteria has been
published.[@sanchezpinto_2024_development;@schlapbach_2024_international]
However, the published code required extensive redactions to protect the
anonymity of both patients and health care systems providing data, resulting in
code that is not easily reusable.

There is a need for a tool that can apply the Phoenix criteria to any electronic
health record (EHR) data set consistently such that publications reporting on or
using the Phoenix criteria can be compared fairly among each other.

To fill this need we have developed and published a R package, Python module,
and example SQL queries for applying the Phoenix criteria to other data sets.

# Objective

To provide an efficient and consistent way to apply the Phoenix scoring rubric
to new EHR data sets and for all researchers, we developed the `phoenix` R
package, Python module, and example SQL queries.

# Materials and Methods

The Phoenix criteria (@tbl-phoenix-rubric) were developed using a dataset that
included over 3.5 million pediatric encounters from ten hospital systems across
North America, South America, Asia, and Africa. The criteria are applicable to
pediatric patients in both high- and low/middle-resourced environments.  The
development data excluded birth-hospitalizations and patients with gestational
ages less than 37
weeks.[@sanchezpinto_2024_development;@schlapbach_2024_international]

::: {#tbl-phoenix-rubric}

 | Organ System                                                          | 0 Points                         | 1 Point                     | 2 Points                 | 3 Points |
 | :------                                                               | :--                              | :--                         | :--                      | :--      |
 | **Respiratory** (0-3 points)                                          |                                  |                             |                          |          |
 |     Respiratory Support                                               |                                  | Any respiratory support     | IMV[^a]                  | IMV      |
 |     PaO~2~:FiO~2~                                                     | &geq; 400                        | &lt; 400                    | &lt; 200                 | &lt; 100 |
 |     SpO~2~:FiO~2~[^b]                                                 | &geq; 292                        | &lt; 292                    | &lt; 220                 | &lt; 148 |
 | **Cardiovascular** (0-6 points; sum of medications, Lactate, and MAP) |                                  |                             |                          |          |
 |      Systemic Vasoactive Medications[^c]                              | No medications                   | 1 medication                | 2 or more medications    |          |
 |      Lactate[^d] (mmol/L)                                             | &lt; 5                           | 5 &leq; Lactate[^j] &lt; 11 | &geq; 11                 |          |
 |      Age[^e] (months) adjusted MAP[^f] (mmHg)                         |                                  |                             |                          |          |
 |            0 &leq; Age &lt;   1                                       | &geq; 31                         | 17 &leq; MAP &lt; 31        | &lt; 17                  |          |
 |            1 &leq; Age &lt;  12                                       | &geq; 39                         | 25 &leq; MAP &lt; 39        | &lt; 25                  |          |
 |           12 &leq; Age &lt;  24                                       | &geq; 44                         | 31 &leq; MAP &lt; 44        | &lt; 31                  |          |
 |           24 &leq; Age &lt;  60                                       | &geq; 45                         | 32 &leq; MAP &lt; 45        | &lt; 32                  |          |
 |           60 &leq; Age &lt; 144                                       | &geq; 49                         | 36 &leq; MAP &lt; 49        | &lt; 36                  |          |
 |          144 &leq; Age &lt; 216                                       | &geq; 52                         | 38 &leq; MAP &lt; 52        | &lt; 38                  |          |
 | **Coagulation**[^g] (0-2 points; 1 for each lab; max of 2 points)     |                                  |                             |                          |          |
 |      Platelets (1000/&mu;L)                                           | &geq; 100                        | &lt; 100                    |                          |          |
 |      INR                                                              | &leq; 1.3                        | &gt; 1.3                    |                          |          |
 |      D-Dimer (mg/L FEU)                                               | &leq; 2                          | &gt; 2                      |                          |          |
 |      Fibrinogen (mg/dL)                                               | &geq; 100                        | &lt; 100                    |                          |          |
 | **Neurologic**[^h] (0-2 points)                                       |                                  |                             |                          |          |
 |                                                                       | GCS[^i] &geq; 11                 | GCS &leq; 10                | Bilaterally fixed pupils |          |
 | **Endocrine** (0-1 point)                                             |                                  |                             |                          |          |
 |      Blood Glucose (mg/dL)                                            | 50 &leq; Blood Glucose &leq; 150 | &lt; 50; or &gt; 150        |                          |          |
 | **Immunologic** (0-1 point; point from ANC and/or ALC)                |                                  |                             |                          |          |
 |      ANC (cells/mm^3^)                                                | &geq; 500                        | &lt; 500                    |                          |          |
 |      ALC (cells/mm^3^)                                                | &geq; 1000                       | &lt; 1000                   |                          |          |
 | **Renal** (0-1 point)                                                 |                                  |                             |                          |          |
 |      Age[^e] (months) adjusted Creatinine (mg/dL)                     |                                  |                             |                          |          |
 |            0 &leq; Age &lt;   1                                       | &lt; 0.8                         | &geq; 0.8                   |                          |          |
 |            1 &leq; Age &lt;  12                                       | &lt; 0.3                         | &geq; 0.3                   |                          |          |
 |           12 &leq; Age &lt;  24                                       | &lt; 0.4                         | &geq; 0.4                   |                          |          |
 |           24 &leq; Age &lt;  60                                       | &lt; 0.6                         | &geq; 0.6                   |                          |          |
 |           60 &leq; Age &lt; 144                                       | &lt; 0.7                         | &geq; 0.7                   |                          |          |
 |          144 &leq; Age &lt; 216                                       | &lt; 1.0                         | &geq; 1.0                   |                          |          |
 | **Hepatic** (0-1 point; point from total bilirubin and/or ALT)        |                                  |                             |                          |          |
 |      Total Bilirubin (mg/dL)                                          | &lt; 4                           | &geq; 4                     |                          |          |
 |      ALT (IU/L)                                                       | &leq; 102                        | &gt; 102                    |                          |          |

The Organ Dysfunction Scoring for the Phoenix Criteria. The Phoenix sepsis
criteria are based on the Phoenix Sepsis Score, which includes respiratory,
cardiovascular, coagulation, and neurologic dysfunction; Phoenix-8 is based on
those four organ systems plus endocrine, immunologic, renal, and hepatic
dysfunction. Sepsis is defined as a Phoenix Sepsis Score &geq; 2. Septic Shock
is defined as sepsis with at least one cardiovascular point.  Missing data maps
to scores of zero. The limits reported in this table reflect the implementation
of the criteria in software, whereas the comparable published tables report the
criteria from a clinical
perspective.[@sanchezpinto_2024_development;@schlapbach_2024_international] The
two representations of the criteria are consistent in practice.
:::


[^a]: Abbreviations:
  ALC: Absolute lymphocyte count;
  ALT: alanine aminotransferase;
  ANC: Absolute neutrophil count;
  FEU: fibrinogen equivalent units;
  FiO~2~: fraction of inspired oxygen;
  GCS: Glasgow Coma Score;
  IMV: invasive mechanical ventilation;
  INR: International normalized ratio;
  MAP: mean arterial pressure;
  PaO~2~: arterial oxygen pressure;
  SpO~2~: pulse oximetry oxygen saturation;

[^b]: SpO~2~:FiO~2~ is only valid when SpO~2~ &leq; 97.

[^c]: Vasoactive medications: any systemic dose of dobutamine, dopamine, epinephrine, milrinone, norepinephrine, and/or vasopressin.

[^d]: Lactate can be arterial or venous. Reference range 0.5 - 2.2 mmol/L

[^e]: Age: measured in months and is not adjusted for prematurity.

[^f]: MAP - Use measured mean arterial pressure preferentially (invasive arterial if available, or non-invasive oscillometric), alternatively use the calculation diastolic + (systolic - diastolic) / 3

[^g]: Coagulation variable reference ranges: platelets, 150-450 103/μL; D-dimer, < 0.5 mg/L FEU; fibrinogen, 180-410 mg/dL. International normalized ratio reference range is based on local reference prothrombin time.

[^h]: Neurologic dysfunction scoring was pragmatically validated in both sedated and on sedated patients and those with and without IMV.

[^i]: GCS measures level of consciousness based on verbal, eye, and motor response. Values are integers from 3 to 15 with higher scores indicating better neurologic function.

[^j]: The verbosity of this table is greater than in the tables in the original source publications.[@sanchezpinto_2024_development;@schlapbach_2024_international] The inequalities reported in this table, and the specific values reported in this table, reflect how the criteria is implemented in software whereas the source publications reported tables consistent with clinical practice.  A couple notable differences.  1 cardiovascular point is reached for a lactate value of “5-10.9 mmol/L” and 2 points for lactate &geq; 11 mmol/L.[@sanchezpinto_2024_development;@schlapbach_2024_international] There is an implication of rounding lactate to one decimal place and assessing the criteria. The software simplifies the work by considering lactate values to be a floating point value that could take on any real value and thus the logic of “5 &leq; lactate &lt; 11” for 1 point.  Additionally, for MAP, the criteria listed in this table is consistent with common clinical practice of interpreting MAP as integer values. The criteria listed in this table is used with the assumption that MAP values are floating point values.

The Phoenix criteria define sepsis as a suspected infection (operationalized as
at least one dose of a systemic anti-microbial medication and at least one
microbiological test ordered within the first 24 hours of a hospital encounter) with
a Phoenix score of at least 2 points.  Additionally, septic shock is defined as
sepsis with at least one point from the cardiovascular dysfunction component of
the Phoenix Sepsis
Score.[@sanchezpinto_2024_development;@schlapbach_2024_international]

Missing data values are mapped to scores of
zero.[@sanchezpinto_2024_development] It was reasonable to assume that for some
laboratory values and metrics missing data indicate no concern and testing
was not ordered. Further, the Phoenix criteria was developed to be useful in
high, medium, and low resource settings where some laboratory values,
medications, and other values might be uncommon or impossible to obtain.  The
`phoenix` R package, Python module, and example SQL queries handle missing
values consistent with the development approach for the Phoenix criteria.

The `phoenix` R package is available from the Comprehensive R
Archive Network (CRAN) (https://cran.r-project.org/package=phoenix) and GitHub
(https://github.com/cu-dbmi-peds/phoenix/).  `phoenix`
was designed to be as light-weight as possible.  There are no dependencies nor
imports save base R.  The R package was the primary focus for development and
will be the focus for this manuscript.  A testing suite for version 1.0.0 of
`phoenix` has 100% code coverage (details on the GitHub page) along with
automatic CRAN checks for Windows, MacOS, and Ubuntu, for the current version
of R, the prior version of R, and the development version of R.

The Python module has been made public via PyPi
(https://pypi.org/project/phoenix-sepsis/) with source code available in the
same GitHub repository as the R package.  A set of tests are built within the
GitHub source code to ensure that the results of the Python module are identical
to the results of the R package.

Lastly, example SQL queries, as understood by SQLite, are also provided in the
GitHub repository and the **Supplemental Examples**.  As with the Python module
there is testing code within the repo to ensure the results of the SQLite
queries are identical to the R package.

Extensive documentation for the R package, Python module, and SQLite queries are
available on the package website https://cu-dbmi-peds.github.io/phoenix/.

<!--

The Phoenix criteria was developed for pediatric patients during non-birth
hospitalizations for patients with a gestational ages &gt; 37 weeks up to 216
months (18 years).

The Phoenix criteria is based on observations within the first 24-hours of a
hospital encounter.  If there are $n$ observations for a patient over
the first 24 hours, then Phoenix is the lower bound of the triangle inequality.
That is, for observations $i = 1, 2, \ldots, n$ in the first 24 hours,
Phoenix Sepsis Total = $\max_i ( resp_i + card_i + coag_i + neur_i)$.
not $\max(resp_i) + \max(card_i) + \max(coag_i) + \max(neur_i)$.  An detailed
example is in Supplement 1.

-->


# Results {#sec-results}

An example data set, `sepsis`, is provided as a `data.frame` within the R
package, as a plain text file in the Python module, and used in the example
SQLite queries. The data consists of `r nrow(sepsis)` synthetic observations of
`r ncol(sepsis)` variables needed by the Phoenix and Phoenix-8 criteria. The
data is lazyloaded in R and is available when the `phoenix` namespace is
attached to the search path, i.e., when `library(phoenix)` is called.

```{r}
#| label: r_data_load
library(phoenix)
dim(sepsis)
names(sepsis)
```

In Python, the example data can be loaded into a pandas DataFrame via
```{python}
#| label: py_data_load
import numpy as np
import pandas as pd
import importlib.resources
import phoenix as phx
path = importlib.resources.files('phoenix')
sepsis = pd.read_csv(path.joinpath('data').joinpath('sepsis.csv'))
print(sepsis.shape)
```

Extensive detail on the synthetic data is available in the R package
documentation, the package website, and in the **Supplementary Examples**.

End users of the Python module and R package will have generally the same
experience.  Both use the same naming conventions and provide the same ten
vectorized functions for applying the Phoenix criteria (@tbl-functions).

::: {#tbl-functions}

 | Phoenix Criteria           | R, Python Function       | R Return       | Python Return    |
 | :----------------          | :-----------------       | :------        | :-----------     |
 | Respiratory dysfunction    | `phoenix_respiratory`    | Integer vector | Numpy array      |
 | Cardiovascular dysfunction | `phoenix_cardiovascular` | Integer vector | Numpy array      |
 | Coagulation dysfunction    | `phoenix_coagulation`    | Integer vector | Numpy array      |
 | Neurologic dysfunction     | `phoenix_neurologic`     | Integer vector | Numpy array      |
 | Endocrine dysfunction      | `phoenix_endocrine`      | Integer vector | Numpy array      |
 | Immunologic dysfunction    | `phoenix_immunologic`    | Integer vector | Numpy array      |
 | Renal dysfunction          | `phoenix_renal`          | Integer vector | Numpy array      |
 | Hepatic dysfunction        | `phoenix_hepatic`        | Integer vector | Numpy array      |
 | Phoenix Criteria           | `phoenix`                | `data.frame`   | Pandas DataFrame |
 | Phoenix-8 Criteria         | `phoenix8`               | `data.frame`   | Pandas DataFrame |

R and Python function names and returns for each of the Phoenix organ
dysfunction scores and total scoring.  The Phoenix criteria is the sum of the
respiratory, cardiovascular, coagulation, and neurologic scores.  Phoenix
sepsis is defined as a total score of two or more points (along with suspected
infection).  Septic shock is sepsis with at least one cardiovascular point.
Phoenix-8 is an extended scoring system and is the sum of all eight organ
systems.

:::


The return of `phoenix()` is a data.frame (R) Pandas DataFrame (Python) with
`r ncol(phoenix())` columns; the respiratory dysfunction score, cardiovascular
dysfunction score, coagulation dysfunction score, neurologic dysfunction score,
total score, and indicator columns for sepsis (total score &geq; 2), and septic
shock (sepsis with &geq; 1 cardiovascular points).  `phoenix8()` returns the
same as `phoenix()` with additional columns for the endocrine, immunologic,
renal, and hepatic dysfunction scores, and the Phoenix-8 total score.  All the
columns are integer valued.

A simple example[@sanchezpinto_2024_development]: a three-year-old presenting
with a fever, tachycardia, and irritability is given broad spectrum antibiotics
and started on a norepinephrine drip due to hypotension (blood pressure 67/32).
A complete blood count (CBC) shows a platelet count of 95 K/&mu;L. Applying the
Phoenix criteria to this patient results in a respiratory score of 0,
cardiovascular score of 2; coagulation score of 1; neurologic score of 0, and a
total score of 3; sepsis=yes because the score is &geq; 2; septic shock=yes
because the cardiovascular score is also &geq; 1. Scoring in R would be done
using:

```{r}
#| echo: true
#| results: markup
library(phoenix)
p <- phoenix(
  vasoactives = 1,
  map         = map(sbp = 67, dbp = 32), # MAP = 2/3 * DBP + 1/3 * SBP,
  platelets   = 95,
  gcs         = 14,
  age         = 36)
str(p)
```

Scoring in Python:
```{python}
#| echo: true
#| results: markup
import phoenix as phx
df = phx.phoenix(
  vasoactives = 1,
  map         = (2/3) * 32 + (1/3) * 67,
  platelets   = 95,
  gcs         = 14,
  age         = 36)
print(df.T)
```

In the above example only the known data need be inputted.  Missing values are
mapped to scores of zero.  That is, Phoenix is based on the explicitly defined
inputs and any missing inputs are implicitly mapping to scores of zero.  This is
consistent with the Phoenix development process.[@sanchezpinto_2024_development]

To apply the Phoenix rubric to a full data set in R:
```{r}
phoenix_scores <-
  phoenix(
    # respiratory
      pf_ratio = pao2 / fio2,
      sf_ratio = ifelse(spo2 <= 97, spo2 / fio2, NA_real_),
      imv = vent,
      other_respiratory_support = as.integer(fio2 > 0.21),
    # cardiovascular
      vasoactives = dobutamine + dopamine + epinephrine +
                    milrinone + norepinephrine + vasopressin,
      lactate = lactate,
      age = age,
      map = dbp + (sbp - dbp)/3,
    # coagulation
      platelets = platelets,
      inr = inr,
      d_dimer = d_dimer,
      fibrinogen = fibrinogen,
    # neurologic
      gcs = gcs_total,
      fixed_pupils = as.integer(pupil == "both-fixed"),
    data = sepsis
  )
str(phoenix_scores)
```

Details on the expected units for inputs as denoted in @tbl-phoenix-rubric are
also provided in the documentation for the R package
```{r}
#| eval: false
?phoenix8
```
and Python module
```{python}
#| eval: false
help(phx.phoenix8)
```

Additional details and examples for each of the eight organ dysfunction scoring
functions, `phoenix()`, and `phoenix8()` methods are in the **Supplemental
Examples**.  The supplement includes examples in R, Python, and SQLite.

# Discussion

The transition to the Phoenix criteria marks a major change in the conceptual
definition of pediatric sepsis.[@jabornisky2024transitioning] Applicable across
differentially resourced settings, the Phoenix criteria should help improve
clinical care and research across the globe.

Additionally, the eponymic R package and Python modules provide researchers a
simple to use tool for consistent and faithful application of the Phoenix
criteria to any applicable data set.

Researchers are encouraged to carefully review the provided documentation for
the package. Some assumptions used by the package are easy to overlook. The
example for using `phoenix()` on a `data.frame` in R has expressions for the
PaO~2~:FiO~2~ ratio, SpO~2~:Fi~2~ ratio, respiratory support, vasoactives
medications, mean arterial pressure (MAP), and fixed pupils.  This example is
provided to be explicit about data assumptions such as the SpO~2~:FiO~2~ ratio
only being valid for SpO~2~ values not exceeding 97.

In practice, we suggest processing the data first such that only a variable name
need be passed as an argument.  This could be particularly useful in the case of
MAP where a hierarchy of values could be used, i.e., invasive MAP readings are
preferable to calculated MAP based on invasive SBP and DBP, and invasive
measurements are preferable to non-invasive blood pressure cuff measurements.

<!--
"When you are doing your research and you call sepsis or you call septic shock
[the patient] must have [the Phoenix] criteria."[@jamapodcast] -- Roberto Jabornisky
-->


## Conclusions

The `phoenix` R package and Python module, meets the objectives of the FAIR
Principles for Research Software (FAIR4RS Principles).[@barker_2022_introducing]
The package and module are intuitive tools for consistently and accurately
applying the Phoenix pediatric sepsis criteria to clinical data sets.

{{< pagebreak >}}

[**Funding Statement:**]{.underline} This work was supported by Eunice Kennedy Shriver National Institute of Child Health and Human Development grant R01HD105939 to TDB and LNSP.

[**Competing Interests Statement:**]{.underline} None.

[**Data Availability Statement:**]{.underline} The `phoenix` R package is freely
available on CRAN at https://cran.r-project.org/package=phoenix.  The `phoenix`
Python module is freely available from PyPi:
https://pypi.org/project/phoenix-sepsis/.
Extensive documentation and example SQLite queries are available online at
https://cu-dbmi-peds.github.io/phoenix/index.html.
This manuscript was written using Quarto version
`r system("quarto --version", intern = TRUE)`
(https://quarto.org/) and `r R.version.string`. All R code, materials,
and dependencies can be found at
https://github.com/cu-dbmi-peds/phoenix_application_note/.

[**Contributorship Statement:**]{.underline} PED designed and developed the
`phoenix` R package, Python module, example SQL code, and drafted the
manuscript. SR provided critical code review and testing.  SR, MR, LNSP, and TDB
all contributed to the conception, design, and interpretation and provided
important feedback. All authors critically reviewed the manuscript for important
intellectual content and approved of the final version of the manuscript.

<!--
[**Acknowledgments:**]{.underline}
-->

{{< pagebreak >}}

# References {.unnumbered}

::: {#refs}
:::

<!--
{{< pagebreak >}}


{{< pagebreak >}}

# Figures {.unnumbered}
-->
